### 2016-05-11 text/Chapter08.tex
Add binding to chapter 2, whitespace, tweaks

@@

\chapter{Implementation}
\label{ch:impl}

In the previous three chapters, we presented the theory of coeffects consisting of type system
and comonadically inspired semantics of two parameterized coeffect calculi. The theory provides
a framework that simplifies the implementation of safe context-aware programming languages. To
support this claim, this chapter presents a prototype implementation of three coeffect languages --
language with implicit parameters and both flat and structural versions of a data-flow language.

The implementation directly follows the thoery presented in the previous three chapters. It
consists of a common framework that provides type checking and translation to a simple functional
target language with comonadically-inspired primitives. Each concrete context-aware language then
adds a domain-specific rule for choosing unique typing derivation (as discussed in
Section~\ref{sec:flat-unique}) together with a domain-specific definition of the
comonadically-inspired primitives that define the runtime semantics (see Section~\ref{sec:semantics-proofs}).

The main goal of the implementation is to show that the theory is practically useful and to present it
in a more practical way. However, we do not intend to build a complete real-world programming language.
For this reason, the implementation is available primarilly as an interactive web-based essay, though
it can be also downloaded and run locally.

\paragraph{Chapter structure and contributions}

\begin{itemize}
\item We discuss how the implementation follows from the theory presented earlier (Section~\ref{sec:impl-theory}).
  This applies to the implementation of the \emph{type checker} and the implementation of the \emph{translation}
  to a simple target langauge that is then iterpereted. We also discuss how the common framework makes it
  easy to implement additional context-aware languages (Section~\ref{sec:impl-theory-ext}).

\item We consider a number of case studies (Section~\ref{sec:impl-case}) that illustrate ointeresting
  aspects of the theories discussed earlier. This includes the typing of lambda abstraction and the
  difference between flat and structural systems (Section~\ref{sec:impl-case-typing}) and the
  comonadically-inspired translation (Section~\ref{sec:impl-case-transl}).

\item We briefly outline some of the interesting technical details of the implementation
  (Section~\ref{sec:impl-tech}). To make the implementation easily accessible, we make it
  available via web and we also implement a limited support for type inference.

\item The implementation is available not just as downloadable code, but also in the format of interactive essay
  (Section~\ref{sec:impl-essay}), which aims to make coeffects accessible to a broader audience.
  We conclude this chapter by discussing the most interesting aspects of the presentation.
\end{itemize}

% ==================================================================================================
%
%    #######
%       #    #    # ######  ####  #####  #   #
%       #    #    # #      #    # #    #  # #
%       #    ###### #####  #    # #    #   #
%       #    #    # #      #    # #####    #
%       #    #    # #      #    # #   #    #
%       #    #    # ######  ####  #    #   #
%
% ==================================================================================================

\section{From theory to implementation}
\label{sec:impl-theory}

The theory discussed so far provides the two key components of the implementation. In
Chapter~\ref{ch:flat}, we discussed the type checking of context-aware programs and
Chapter~\ref{ch:semantics} models the execution of context-aware programs (in terms of
translation and operational semantics). For structural coeffects, the same components are discussed
in Chapter~\ref{ch:structural}. In this section, we discuss how those provide foundation for the
implementation.

\subsection{Type checking}
\label{sec:impl-theory-typing}

(Section~\ref{sec:flat-unique})

\paragraph{Implementation.}
From the presentation in this section, it might appear that resolving the ambiguity related to
lambda abstraction for implicit parameters requires a type system that is quite different from
the core flat coeffect type system shown earlier in Figure~\ref{fig:flat-types}. This is not the
case. As discussed in Chapter~\ref{ch:impl}, the required changes in the implementation are
simpler.

Briefly, the implementation collects constraints on the coeffects and then finds the smallest sets
of implicit parameters to satisfy the constraints. We still need to track implicit parameters in
scope $\cclrd{\Delta}$, but the rest of the (\emph{abs}) rule from the implementation is close
to the one from Figure~\ref{fig:flat-types}:
%
\begin{equation*}
\tyrule{abs}
  {\coctx{\Gamma, x\!:\!\tau_1;\cclrd{\Delta}}{\cclrd{t}} \vdash e : \tau_2 ~|~ C}
  {\coctx{\Gamma;\cclrd{\Delta}}{\cclrd{r}} \vdash \lambda x\!:\!\tau_1.e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 ~|~
    C\cup\{\cclrd{t}=\cclrd{r} \,\czip\, \cclrd{s}, \cclrd{r}=\cclrd{\Delta} \}}
\end{equation*}
%
Given a typing derivation for the body that produced constraints $C$, we generate an additional
constraint that restricts $\cclrd{r}$ (declaration-site demands) to those available in the
current static scope $\cclrd{\Delta}$. The constraint satisfaction algorithm then finds
the minimal set $\cclrd{s}$ which is $\cclrd{t}\cclrd{\setminus \Delta}$.

\paragraph{Implementation.}
As with implicit parameters, the implementation (discussed in Chapter~\ref{ch:impl}) does not
require changing the typing (\emph{abs}) rule of the flat coeffect system. The typing specified
by (\emph{idabs}) can be easily obtained by generating additional constraints:
%
\begin{equation*}
\tyrule{abs}
  {\coctx{\Gamma, x\!:\!\tau_1}{\cclrd{t}} \vdash e : \tau_2 ~|~ C}
  {\coctx{\Gamma}{\cclrd{s}} \vdash \lambda x\!:\!\tau_1.e : \tau_1 \xrightarrow{\cclrd{t}} \tau_2 ~|~
    C\cup\{\cclrd{t}=\cclrd{r} \,\czip\, \cclrd{s}, \cclrd{r}=\cclrd{t}, \cclrd{s}=\cclrd{t} \}}
\end{equation*}
%
Here, the two additional constraints restrict both $\cclrd{r}$ and $\cclrd{s}$ to be equal to the
coeffect of the body $\cclrd{t}$ and so the only possible resolution is the one specified by
(\emph{idabs}).

Type checking
 * DSL specific constraint resolution
 * easier for structural

\subsection{Translation}
\label{sec:impl-theory-transl}

Translation ~ Semantics

\subsection{Extensions}
\label{sec:impl-theory-ext}

It is really easy to add another DSL

\section{Case studies}
\label{sec:impl-case}

\subsection{Typing}
\label{sec:impl-case-typing}

let both =
  let ?fst = 100 in
  fun trd -> ?fst + ?snd + trd in
let ?fst = 200 in
both 1

num -{ ?snd:num }-> num

let struct x y = x + prev y in
struct

num -{ 1 }-> num -{ 1 }-> num

let flat x y = x + prev y in
flat

num -{ 0 }-> num -{ 1 }-> num

\subsection{Translation}
\label{sec:impl-case-transl}

let ?param = 10 in
fun x -> ?param + ?other

let (ctx2, ctx3) = split (duplicate finput) in
let ctx1 = letimpl ?param (ctx2, 10) in
fun x ->
  let ctx4 = merge (x, ctx1) in
  let (ctx5, ctx6) = split (duplicate ctx4) in
  lookup ?param ctx5 + lookup ?other ctx6

fun x -> (fun v -> prev v) (prev x)

fun x ->
  let ctx1 = merge (x, sinput) in
  (fun v ->
    let ctx3 = merge (v, choose_0 ctx1) in
    counit (prev ctx3)) (cobind (fun ctx2 ->
      counit (prev ctx2)) (choose_1 ctx1))

\section{Technical details}
\label{sec:impl-tech}

1. F# + JavaScript
2. Single AST and parsing
3. Type checking and constraint generation
4. Constraint resolution

\section{Interactive essay}
\label{sec:impl-essay}

>>>>

\chapter{Related and future work}
\label{ch:further}

%===================================================================================================

The main goal of this thesis is to provide a \emph{unified} calculus for tracking context
dependence. We have not achieved this goal yet. In Chapter~\ref{ch:applications}, we identified
two kinds of contextual properties that we further covered separately. Flat coeffects in
Chapter~\ref{ch:flat} track whole-context properties, while structural coeffects, covered in
Chapter~\ref{ch:structural}, track per-variable properties. In this chapter, we unify the two
notions.

Although the results presented in this thesis are mainly of a theoretical nature, we indeed
believe that coeffects should be integrated in main-stream programming languages. This is addressed
in the second part of the chapter, where we outline possible approach for a practical implementation
of coeffect systems. Finally, we conclude the chapter with a brief discussion of an alternative
approach to coeffect system, based on the meta-language style.

\paragraph{Chapter structure and contributions}
\begin{itemize}
\item We start by unifying the two kinds of coeffects discussed so far. In
  Section~\ref{sec:further-unified}, we introduce the \emph{unified coeffect calculus}, which
  generalizes flat and structural systems and can be instantiated to track both flat and structural
  properties.

\item We outline one possible approach for practical implementations of coeffects
  (Section~\ref{sec:further-impl}). The technique is based on embedding comonadic computations
  via a lightweight syntactic extension and using rich type systems for capturing the structure of
  coeffect algebra.

\item Finally, we discuss an alternative approach to defining coeffect systems based on the
  meta-language style, which highlights the relationship between our work and related work arising
  from modal logics such as CMTT \cite{logic-cmtt} (Section~\ref{sec:further-meta}).
\end{itemize}

\noindent
The unified coeffect calculus presented in the first part of the chapter is a novel result that
completes the theory developed in this thesis. The next two sections present important future and
related work, respectively.

@@

\section{The unified coeffect calculus}
\label{sec:further-unified}

The type systems of the flat coeffect calculus (Figure~\ref{fig:flat-types}) and the structural
coeffect calculus (Figure~\ref{fig:struct-types}) differ in a number of ways. Understanding the
differences is the key to reconciling the two systems:

\begin{itemize}
\item The structural coeffect calculus contains explicit rules for context manipulation
  (weakening, contraction, exchange). In the flat coeffect calculus, these rules are not defined
  explicitly, but are admissible.

\item In the structural coeffect calculus, the variable context is treated as a vector
  and is annotated with a vector of (scalar) coeffects. In the flat coeffect calculus,
  the variable context is a finite partial function (from names to types) and is annotated with
  a single (scalar) coeffect.

\item In the flat coeffect calculus, we distinguish between \emph{splitting} context requirements
  and \emph{merging} of context requirements ($\cpar$ and $\czip$, respectively). In the structural
  coeffect calculus, the operations (which model appending and splitting vectors) are invertible
  and so the structural coeffect algebra uses the same tensor product $\atimes$.
\end{itemize}

\noindent
In the unified calculus presented in this section, we address the three differences as follows.
We include explicit rules for context manipulation in the calculus; in systems that arise from flat
calculi, the structural rules do not change coeffects and can thus be applied freely. We generalize
the structure of coeffect annotations using the notion of a ``container'' which can be specialized
to obtain a single annotation or a vector of annotations. This could potentially be used to capture
other structures such as trees in bunched typing \cite{substruct-bunched}. Finally, we distinguish
between splitting and merging of context requirements (using the notation $\apar$ and $\azip$,
respectively). For structural coeffect calculi, the two operators coincide, but for flat
coeffect calculi, they provide the needed flexibility.

\subsection{Shapes and containers}
Our generalization of coeffect structure using a \emph{coeffect container} is based on containers of
Abbott et al. \cite{types-containers}. Interestingly, containers have later been linked
to comonads by Ahman et al. \cite{comonads-containers}. Intuitively a container describes data
types such as lists, trees or streams. A container is formed by shapes (\eg~lengths of
lists)\footnote{Shapes can also be intuitively thought of as sizes, but this is not fully precise
as some containers may have multiple shapes of the same length, \eg~differently balanced trees.}.
For every shape, we can obtain a set of positions in the container (\eg~offsets in a list of a
specified length). More formally:

\begin{definition}
A \emph{container} consists of a pair $(S,P)$, usually written as $S \triangleleft P$.
Here, $S$ is a set of shapes and $P$ is a shape-indexed family $P : S \rightarrow \ident{Set}$
of positions.
\end{definition}

\noindent
Well-known examples of containers include lists, non-empty lists, (un\-boun\-ded) streams,
trees and the singleton data type (which contains exactly one element). Containers relevant
to our work are lists and singleton data types:

\begin{itemize}
\item The container representing lists is given by $S \triangleleft P$ where
  shapes are integers $S = \ident{Nat}$ (lengths of a list). The set of positions for a
  given length $n$ is the set of indices $P(n)= \{ 1 \ldots n \}$.

\item The container representing the singleton data type is given by $S \triangleleft P$ where
  shapes are given by a singleton set $S = \{ \ast \}$ and the set of positions for the
  shape $\ast$ contains exactly one position. To follow the intuition based on offset or
  index, we write the single position as $0$, that is $P(\ast) = \{ 0 \}$.
\end{itemize}

\noindent
In the unified coeffect calculus, the structure of coeffect annotations is defined by a
container with additional operations (discussed later) that links it with the free-variable
context $\Gamma$.

\subsection{Structure of coeffects}
In the structural coeffect calculus, the coeffect annotation was formed by a vector of coeffect
scalars. The annotations in the unified coeffect calculus are similar, but a \emph{vector} is
replaced with a more general \emph{container}. The primitive coeffect annotations in the unified
calculus are formed of \emph{coeffect scalars}. The coeffect scalars are equipped with the same
structure as in structural coeffect calculus (Definition~\ref{def:structural-scalar}). In this
section, we refer to it as \emph{\cclrd{unified coeffect scalar}} (and we repeat the definition
below). Then we define \emph{\sclrd{unified coeffect containers}} which determines how coeffect
scalar values are attached to the free-variable context. Finally, we define the
\emph{\aclrd{unified coeffect algebra}} which consists of shape-indexed coeffect scalar values.

As in the structural coeffect calculus, the contexts in the unified calculus are annotated with
shape-indexed coeffects, written as $\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash e : \tau$;
functions take just a single input parameter and so are annotated with scalar coeffect values
$\sigma \xrightarrow{\cclrd{r}} \tau$.

\paragraph{Coeffect scalars.}
The following definition of the coeffect scalar structure repeats the
Definition~\ref{def:structural-scalar} from the previous chapter.

\begin{definition}
A \emph{\cclrd{unified coeffect scalar}} $(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ is a set
$\C$ together with elements $\cunit, \czero \in \C$, relation $\cleq$ and binary operations
$\cseq, \cpar$ such that $(\C, \cseq, \cunit)$ and $(\C, \cpar, \czero)$ are monoids and
$(\C, \cleq)$ is a pre-order. That is, for all $r,s,t\in \C$:
%
\begin{equation*}
\begin{array}{ccr}
r \;\cseq\; (s \;\cseq\; t) = (r \;\cseq\; s) \;\cseq\; t  &
\cunit \;\cseq\; r = r = r \;\cseq\; \cunit &
\textnormal{(monoid)}
\\
r \;\cpar\; (s \;\cpar\; t) = (r \;\cpar\; s) \;\cpar\; t &
\czero \;\cpar\; r = r = r \;\cpar\; \czero &
\textnormal{(monoid)}
\\
\textnormal{if}~~r\; \cleq\; s ~~\textnormal{and}~~s\; \cleq\; t~~\textnormal{then}~~r\; \cleq\; t&
t\; \cleq\; t &
\textnormal{(pre-order)}
\\
\end{array}
\end{equation*}
%
The following distributivity axioms hold:
\begin{align*}
\quad (\cclrd{r}\, \cpar\, \cclrd{s}) \;\cseq\; \cclrd{t} & = (\cclrd{r} \,\cseq\, \cclrd{t}) \;\cpar\; (\cclrd{s}\, \cseq\, \cclrd{t}) \\
\quad \cclrd{t} \;\cseq\; (\cclrd{r}\, \cpar\, \cclrd{s}) & = (\cclrd{t} \,\cseq\, \cclrd{r}) \;\cpar\; (\cclrd{t}\, \cseq\, \cclrd{s})
\end{align*}
%
In addition, we require the following two properties of $\czero, \cpar$ and $\cleq$:
\begin{equation*}
\begin{array}{l}
 \quad \cclrd{r}\,\cleq\,\cclrd{r'} \; \Rightarrow \; \forall\cclrd{s}.\;(\cclrd{r}\,\cseq\,\cclrd{s})\;\cleq\;(\cclrd{r'}\,\cseq\,\cclrd{s}) \\
 \quad \czero\;\cleq\;(\czero\,\cseq\,\cclrd{r})
\end{array}
\end{equation*}
\end{definition}

\noindent
As previously, the monoid $(\C,\cseq,\cunit)$ models sequential composition; the laws guarantee
an underlying category structure; $\cunit$ and $\czero$ represent an accessed and unused variable,
respectively.

The $\cpar$ operation models combining of context requirements arising from multiple parts of a
program. The meaning depends on the coeffect container. The operation can either combine requirements
of individual variables (structural coeffects) or requirements attached to the whole context of
multiple sub-exp\-ressions (flat coeffects).

\paragraph{Coeffect containers.}
The coeffect container is a container that determines how are scalar coeffects attached to
free-variable contexts. In addition to a container $\SHP$ formed by shapes and shape-indexed positions,
the coeffect container provides a mapping that returns the shape corresponding to a free-variable context.

The mapping between the shape of the variable context and the shape of the coeffect annotation
is not necessarily bijective. For example, coeffect annotations in flat systems have just a single
shape $\SH = \{ \ast \}$.

In the coeffect judgment $\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash e : \tau$, the coeffect annotation
$\aclrd{\textbf{r}}$ is drawn from the set of coeffect scalars $\C$ indexed by the shape corresponding
to $\Gamma$. We write $\sclrd{s} = \slen{\Gamma}$ for the shape corresponding to $\Gamma$. The operation
$\SP(\sclrd{s})$ returns a \emph{set} of positions and so we can write $\aclrd{\textbf{r}} \in
  \SP(\sclrd{s}) \rightarrow \C$  as a mapping from positions (defined by the shape) to scalar coeffects.
We usually write this as the exponent $\aclrd{\textbf{r}} \in \C^{\SP(\sclrd{s})}$.
The coeffect container is also equipped with an operation that appends shapes (when we concatenate
variable contexts) and two special shapes in $\SH$ representing the empty context and the singleton
context.

\begin{definition}
A \emph{\sclrd{coeffect container}} structure $(\SHP, \stimes, \sempty, \sunit, \slen{-})$
comprises a container $\SHP$ with a binary operation $\stimes$ on $\SH$ for appending shapes, a
mapping from free-variable contexts to shapes $\slen{\Gamma} \in \SH$, and elements $\sempty,\sunit \in
\SH$ such that $(\SH, \stimes, \sempty)$ is a monoid.

The elements $\sempty$ and $\sunit$ represent the shapes of the empty and the singleton free-variable
context, respectively. The $\stimes$ operation corresponds to concatenation of free-variable contexts.
Given $\Gamma_1$ and $\Gamma_2$ such that $\sclrd{s_1}=\slen{\Gamma_1}, \sclrd{s_2}=\slen{\Gamma_2}$,
we require that $\sclrd{s_1}\,\stimes\,\sclrd{s_2}=\slen{\Gamma_1, \Gamma_2}$.
\end{definition}

\noindent
As discussed earlier, we use two kinds of coeffect containers that describe the structure of vectors
(for structural coeffects) and the shape of trivial singleton container (for flat coeffects):

\begin{example}
\label{ex:further-struct-shape}
Structural coeffect container is defined as $\sclrd{(\SHP, {|}\textnormal{--}{|}, +, 0, 1)}$
where $\SH=\mathbb{N}$ and $\SP(\sclrd{n})=\{ 1 \ldots n \}$. The shape mapping ${|}\Gamma{|}$ returns the
number of variables in $\Gamma$. Empty and singleton contexts are annotated with $0$ and $1$,
respectively, and shapes of combined contexts are added so that ${|}\Gamma_1, \Gamma_2{|} =
{|}\Gamma_1{|} + {|}\Gamma_2{|}$.

Therefore, a coeffect annotation is a \emph{vector}
$\aclrd{\textbf{r}} \in \C^{\SP(\sclrd{n})}$ and assigns a coeffect scalar $\aclrd{\textbf{r}}(i) \in \C$
for each position (corresponding to a variable $x_i$ in the context).
\end{example}

\begin{example}
\label{ex:further-flat-shape}
Flat coeffect container is defined as $\sclrd{(\SHP, {|}-{|}, \diamond, \sflat, \sflat)}$.
The container is defined as a singleton data type $\SH = \{\sflat \}$ and $\SP(\sflat) = \{ 0 \}$
with a constant function ${|}\Gamma{|}=\sflat$ and a trivial operation $\sflat\,\diamond\,\sflat = \sflat$.

That is, there is a single shape
$\sflat$ with a single position and all free-variable contexts have the same singleton shape.
Therefore, a coeffect annotation is drawn from $\C^{\{\sflat\}}$ which is isomorphic to $\C$
and so a coeffect scalar $\cclrd{r}\in \C$ is associated with every free-variable context.
\end{example}

\begin{example}
Similarly to the previous example, we can also define a coeffect container with \emph{no} positions,
\ie~$\sclrd{(\SHP, {|}-{|}, \diamond, \sflat, \sflat)}$ where $\SH = \{\sflat \}$, $\SP(\sflat) = \emptyset$,
${|}\Gamma{|}=\sflat$ and $\sflat\,\diamond\,\sflat = \sflat$.
This reduces our system to the simply-typed $\lambda$-calculus with no context annotations, because
$\SP(\sflat)=\emptyset$ and so coeffect annotations would be from the empty set $\C^\emptyset$.
\end{example}

\paragraph{Unified coeffect algebra.}
The unified coeffect calculus annotates typing judgments with shape-indexed coeffect annotations.
The \emph{\aclrd{unified coeffect algebra}} combines a coeffect scalar and a coeffect container
to define shape-indexed coeffects and operations for manipulating these.

The definition here reconciles the third point discussed in Section~\ref{sec:further-unified} --
the fact that flat coeffects use separate operations for splitting and merging ($\cpar$ and $\czip$)
while structural coeffects use the tensor $\atimes$. In the unified calculus, we use two operators
that can, however, coincide.

\begin{definition}
Given a \cclrd{unified coeffect scalar} $(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ and a
\sclrd{coeffect container} $(\SHP, \slen{-}, \stimes, \sempty, \sunit)$ a \emph{\aclrd{unified
coeffect algebra}} extends the two structures with $(\azip, \apar, \azero)$ where $\azero \in \C^{\SP(\sempty)}$
is a coeffect annotation for the empty context and $\azip, \apar$ are families of operations that
combine coeffect annotations indexed by shapes. That is $\forall \sclrd{n},\sclrd{m} \in \SH$:
%
\begin{equation*}
\apar_{\sclrd{m},\sclrd{n}}, \azip_{\sclrd{m},\sclrd{n}} ~:~ \C^{\SP \sclrd{m}} \times \C^{\SP \sclrd{n}} \rightarrow \C^{\SP (\sclrd{m}\,\stimes\,\sclrd{n})}
\end{equation*}
\end{definition}

\noindent
A coeffect algebra induces the following three additional operations:
%
\begin{equation*}
\begin{array}{rclcrcl}
 \alift{-} &\narrow{:}& \C \rightarrow \C^{\SP(\sunit)}  \\
 \alift{\cclrd{x}} &\narrow{=}& \lambda \_ . \cclrd{x}   \\
\\[-0.5em]
 \aseq_{\sclrd{m}} &\narrow{:}&  \C \times \C^{\SP(\sclrd{m})} \rightarrow \C^{\SP(\sclrd{m})}   \\
 \cclrd{r}\, \aseq\, \aclrd{\textbf{s}} &\narrow{=}& \lambda \sclrd{s} . \cclrd{r}\, \cseq\, \aclrd{\textbf{s}}(\sclrd{s}) \\
\\[-0.5em]
 \alen{-} &\narrow{:}& \C^{\SP(\sclrd{m})} \rightarrow \sclrd{m}\\
 \alen{\aclrd{\textbf{r}}} &\narrow{=}& \sclrd{m}
\end{array}
\end{equation*}
%
The $\alift{-}$ operation lifts a scalar coeffect to a shape-indexed coeffect that is indexed by
the shape of a singleton context. The $\aseq_{\sclrd{m}}$ operation is a left multiplication of a
vector by a scalar. As we always use bold face for vectors and ordinary face for scalars (as well
as a distinct colour), using the same symbol is not ambiguous. We also tend to omit the subscript
$\sclrd{m}$ and write~$\aseq$.

Finally, we define $\alen{-}$ as an operation that returns the shape of a given shape-indexed
coeffect. The only purpose is to simplify notation, as we tend to avoid subscripts, but often need
to specify that shapes of variable context and coeffect match,
\eg~$\alen{\aclrd{\textbf{r}}} = \slen{\Gamma}$.

\paragraph{Splitting and merging coeffects.}
The operators $\apar$ and $\azip$ combine shape-indexed coeffects associated with two contexts.
For example, assume we have $\Gamma_1$ and $\Gamma_2$ with coeffects $\aclrd{\textbf{r}} \in
\C^{\SP(\sclrd{m})}$ and $\aclrd{\textbf{s}} \in \C^{\SP(\sclrd{n})}$. In the structural system, the
context shapes $\sclrd{m}, \sclrd{n}$ denote the number of variables in the two contexts. The
combined context $\Gamma_1, \Gamma_2$ has a shape $\sclrd{m}\,\stimes\,\sclrd{n}$ and the combined
coeffects $\aclrd{\textbf{r}} \,\azip\, \aclrd{\textbf{s}}, \aclrd{\textbf{r}} \,\apar\, \aclrd{\textbf{s}}
\in \C^{\SP(\sclrd{m} \stimes \sclrd{n})}$ are indexed by that shape.

For structural coeffect systems such as bounded reuse, both $\apar$ and $\azip$ are
just the tensor product $\times$ of vectors. For flat coeffect systems, the operations
can be defined independently, letting $\azip=\czip$ and $\apar=\cpar$.

The difference between $\azip$ and $\apar$ is clarified by the semantics (Sec\-tion~\ref{sec:further-semantics}),
where $\aclrd{\textbf{r}}\,\azip\,\aclrd{\textbf{s}}$ is an annotation of the \emph{codomain} of
a morphism that merges the capabilities provided by two contexts (in the syntactic
reading, splits the context requirements), while $\aclrd{\textbf{r}}\,\apar\,\aclrd{\textbf{s}}$
is an annotation of the \emph{domain} of a morphism that splits the capabilities of a single
context into two parts (in the syntactic reading, merges their context requirements).
Syntactically, this means that we always use $\azip$ in the rule \emph{assumptions} and
$\apar$ in \emph{conclusions}.

% --------------------------------------------------------------------------------------------------

\newcommand{\tystmt}[2]{ (\text{\footnotesize{#1}})~~{#2} }
\newcommand{\ctxtrans}[3]{ #2 \rightsquigarrow #1, #3 }
\newcommand{\ctxtransnl}[3]{ \begin{array}{l} #2 \rightsquigarrow\\ #1, #3 \end{array}}

\begin{figure}[t]

{\small a.) Syntax-driven typing rules:}
\begin{equation*}
\tyrule{var}
  { }
  {\coctx{x\!:\!\tau}{\alift{\cunit}} \vdash x : \tau }
\end{equation*}
\begin{equation*}
\tyrule{const}
  {c \!:\! \tau \in \Delta}
  {\coctx{()}{\azero} \vdash c : \tau }
\end{equation*}
\begin{equation*}
\tyrule{app}
  {\coctx{\Gamma_1}{\aclrd{\textbf{r}}} \vdash e_1 : \tau_1 \xrightarrow{\cclrd{t}} \tau_2 &
   \coctx{\Gamma_2}{\aclrd{\textbf{s}}} \vdash e_2 : \tau_1 }
  {\coctx{\Gamma_1, \Gamma_2}{\aclrd{\textbf{r}\;\apar\;(\cclrd{t} \,\aseq\, \aclrd{\textbf{s}})}} \vdash e_1~e_2 : \tau_2}
\end{equation*}
\begin{equation*}
\tyrule{abs}
  {\coctx{\Gamma, x\!:\!\tau_1}{\aclrd{\textbf{r}}\,\azip\,\alift{\cclrd{s}}} \vdash e : \tau_2}
  {\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash \lambda x.e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 }
\end{equation*}
\begin{equation*}
\tyrule{let}
  { \coctx{\Gamma_1}{\aclrd{\textbf{r}}} \vdash e_1 : \tau_1 &
    \coctx{\Gamma_2, x\!:\!\tau_1}{\aclrd{\textbf{s}}\;\azip\;\alift{\cclrd{t}}} \vdash e_2 : \tau_2}
  {\coctx{\Gamma_1, \Gamma_2}{(\cclrd{t} \,\aseq\, \aclrd{\textbf{r}}) \;\apar\; \aclrd{\textbf{s}}} \vdash \kvd{let}~x=e_1~\kvd{in}~e_2 : \tau_2 }
\end{equation*}

\vspace{1.5em}
{\small b.) Structural rules for context manipulation:}
\begin{equation*}
\tyrule{sub}
  {\coctx{\Gamma_1,x\!:\!\tau_1,\Gamma_2}{\aclrd{\textbf{r}} \azip \alift{\cclrd{s'}} \azip \aclrd{\textbf{q}}} \vdash e : \tau}
  {\coctx{\Gamma_1,x\!:\!\tau_1,\Gamma_2}{\aclrd{\textbf{r}} \apar \alift{\cclrd{s}} \apar \aclrd{\textbf{q}}} \vdash e : \tau}
\quad
(\cclrd{s'} \cleq \cclrd{s})
\end{equation*}
\begin{equation*}
\tyrule{weak}
  {\coctx{\Gamma}{ \aclrd{\textbf{r}} } \vdash e : \tau}
  {\coctx{\Gamma,x \!:\! \tau_1}{\aclrd{\textbf{r}} \apar \alift{ \czero }} \vdash e : \tau}
\end{equation*}
\begin{equation*}
\tyrule{exch}
  {\coctx{\Gamma_1,x\!:\!\tau_1,y\!:\!\tau_2,\Gamma_2}{\aclrd{\textbf{r}} \azip \alift{\cclrd{s}} \azip \alift{\cclrd{t}} \azip \aclrd{\textbf{q}}} \vdash e : \tau}
  {\coctx{\Gamma_1,y\!:\!\tau_2,x\!:\!\tau_1,\Gamma_2}{\aclrd{\textbf{r}} \apar \alift{\cclrd{t}} \apar \alift{\cclrd{s}} \apar \aclrd{\textbf{q}}} \vdash e : \tau}
\quad
\begin{array}{l}
 \slen{\Gamma_1} = \alen{\aclrd{\textbf{r}}}\\[-0.25em]
 \slen{\Gamma_2} = \alen{\aclrd{\textbf{s}}}
\end{array}
\end{equation*}
\begin{equation*}
\hspace{2.5em}
\tyrule{contr}
  {\coctx{\Gamma_1,y\!:\!\tau_1,z\!:\!\tau_1,\Gamma_2}{\aclrd{\textbf{r}} \azip \alift{\cclrd{s}} \azip \alift{\cclrd{t}} \azip \aclrd{\textbf{q}}} \vdash e : \tau}
  {\coctx{\Gamma_1,x\!:\!\tau_1,\Gamma_2}{\aclrd{\textbf{r}} \apar \alift{\cclrd{s} \,\cpar\, \cclrd{t}} \apar \aclrd{\textbf{q}}} \vdash \subst{e}{z,y}{x} : \tau}
~
\begin{array}{l}
 \slen{\Gamma_1} = \alen{\aclrd{\textbf{r}}}\\[-0.25em]
 \slen{\Gamma_2} = \alen{\aclrd{\textbf{s}}}
\end{array}
\end{equation*}

\vspace{0.5em}
\figcaption{Type system for the unified coeffect calculus}
\label{fig:unif-types}
\end{figure}

% --------------------------------------------------------------------------------------------------

\subsection{Unified coeffect type system}
\label{sec:further-types}
The unified coeffect system in Figure~\ref{fig:unif-types} resembles the structural type system
shown in Figure~\ref{fig:struct-types}. Rather than explaining the rules one-by-one, we focus on
how the unified system differs from the structural system.

The type system for the unified coeffect calculus is parameterized by a coeffect scalar
$(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ together with a coeffect algebra $(\azip, \apar, \azero)$
and the derived constructs $\alift{-}$, $\alen{-}$ and $\aseq$.
As in the structural system, free-variable contexts $\Gamma$ are treated as vectors modulo duplicate
use of variables -- the associativity is built-in. The order of variables matters, but can be changed
using the exchange rule. The context annotations $\aclrd{\textbf{r}}, \aclrd{\textbf{s}}, \aclrd{\textbf{t}}$
are shape-indexed coeffects (rather than simple vectors as before). As before, functions are
annotated with coeffects scalars.

\paragraph{Syntax-driven rules.}
The (\emph{var}) rule is syntactically the same as in the structural system, but it should be read
differently. The $\alift{-}$ operation does not create a \emph{vector}, but a \emph{shape-indexed
coeffect} that returns $\cunit$ for all positions in the singleton shape $\sunit$.
The (\emph{const}) rule annotates empty context with a special annotation of the shape $\sempty$.

In a structural system, the two annotations correspond to a singleton and empty vector, respectively.
However, for a singleton shape with one position, the annotations are equivalent to annotating
variables with a scalar $\cunit$ and constants with a scalar $\czero$.

In the (\emph{app}), (\emph{abs}) and (\emph{let}) rules, the only change from the structural
system is that the vector concatenation $\atimes$ is now replaced with context splitting/merging
of the unified coeffect algebra. As already mentioned, we use splitting of context requirements $\azip$
in rule \emph{assumptions} and merging of context requirements $\apar$ in rule \emph{conclusions}.

Note that we use the terms \emph{merging} and \emph{splitting} in the syntactic (top-down) sense. As
discussed in Chapter~\ref{ch:flat}, Section~\ref{sec:flat-calculus-undestanding}, we can also read
the rules in semantic (bottom-up) sense, in which case assumptions merge available contextual
information and conclusions split available contextual information.

\paragraph{Structural rules.}
The merging/splitting operations in the structural rules are changed in the same way as in the
rules above. It is also worth noting that structural coeffect system used vectors of multiple
elements. For example, $\alift{\cclrd{s},\cclrd{t}}$ and $\alift{\cclrd{t},\cclrd{s}}$ in the (\emph{exch})
rule denoted two-element vectors. In the unified system, this is replaced with merging/splitting
of two lifted scalars: $\alift{\cclrd{s}} \azip \alift{\cclrd{t}}$ and $\alift{\cclrd{t}} \apar \alift{\cclrd{s}}$.

In structural systems, the two notations mean the same thing -- we are simply concatenating or splitting two
singleton vectors. However, this generalization allows us to capture flat coeffect systems as well.
The lifting operation in flat systems simply returns the lifted scalar and operators $\azip$ and
$\apar$ correspond to operations on coeffect scalars. As discussed in Section~\ref{sec:further-flat},
thanks to the properties of coeffect scalars, contraction, weakening and exchange that do not
affect the coeffect annotation are admissible for all flat systems embedded in the unified
coeffect calculus.

% --------------------------------------------------------------------------------------------------

\subsection{Structural coeffects}
\label{sec:further-structural}

The unified coeffect system uses a general notion of context shape, but it has been designed with
structural and flat systems in mind. In this and the next section, we show how it captures
the two coeffect systems from Chapter~\ref{ch:flat} and Chapter~\ref{ch:structural}.

The unified calculus closely resembles the structural system and so using it to model structural
systems is easy -- given a coeffect scalar, we use the coeffect container that describes a \emph{vector}
of annotations (Example~\ref{ex:further-struct-shape}) and define a coeffect algebra formed by
a vector (free monoid) of scalars.

\begin{definition}
\label{def:further-struct}
Given a coeffect scalar $(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ a \emph{structural coeffect system}
is defined by:
